---
title: "Barrow climate data"
output:
  html_document:
    df_print: paged
---

This is a small analysis of Barrow daily weather summaries sourced from NOAA (https://www.ncdc.noaa.gov/cdo-web/datasets). 

I was inspired to look at Barrow weather data by reading in the Washington Post about how the data there was setting off outlier detection climate models (https://www.washingtonpost.com/news/capital-weather-gang/wp/2017/12/12/barrow-is-warming-so-fast-algorithms-removed-the-data-because-it-seemed-unreal/?utm_term=.7a5b322526b4) 
and reading this climate.gov article about rising average temperatures for October (https://www.climate.gov/news-features/understanding-climate/barrow-alaska-climate-change-action)

```{r}
library(ggplot2)
library(plyr)
library(forecast)
library(ggmap)

# read NOAA daily summary data for Barrow
barrow <- read.csv("~/Documents/Climate-Weather/barrowAK_daily.csv", stringsAsFactors=FALSE)
barrow = barrow[,c('STATION','NAME','LATITUDE','LONGITUDE','ELEVATION','DATE'
                   ,'AWND','PGTM','PRCP','SNOW','TMIN','TAVG','TMAX')]
str(barrow)
range(barrow$DATE)
```

This data is from the Barrow airport weather station and spans from Sept 1901 to Dec 2017. 

Let's look at where Barrow is on map of Alaska with the Arctic Circle (66.33 degrees latitude) shown:

```{r}
barrow_location = barrow[1,c('LONGITUDE','LATITUDE')]
barrow_map = get_map(location='alaska',zoom=4)
ggmap(barrow_map)+geom_point(data=barrow,aes(x=LONGITUDE,y=LATITUDE),shape=1,color='red',size=3)+
  geom_hline(lty=2,aes(yintercept=66.33))
```


There are gaps in the earlier years of the data:

```{r}
barrow$obs_date = as.Date(barrow$DATE)
barrow$year = as.numeric(unlist(lapply(barrow$DATE,FUN=function(s){strsplit(s,split='-')[[1]][1]})))
barrow$month = as.numeric(unlist(lapply(barrow$DATE,FUN=function(s){strsplit(s,split='-')[[1]][2]})))
barrow$dayofmonth = as.numeric(unlist(lapply(barrow$DATE,FUN=function(s){strsplit(s,split='-')[[1]][3]})))

ggplot(barrow,aes(obs_date,TMAX))+geom_point(size=0.1)+theme_bw()
```

From October 1920 on is available without any major gaps in the data.

It may be interesting to compare some of the earliest and latest years in the data

```{r}
ggplot(subset(barrow,DATE<'1905-01-01'|DATE>'2015-01-01'),aes(obs_date,TMAX))+geom_point(size=0.1)+geom_line()+theme_bw()+facet_grid(~I(DATE>'1915-01-01'),scales='free_x')
```

Following the climate.gov article, let's look at monthly averages over time starting in Oct 1920:

```{r fig4, fig.height = 10, fig.width = 12}
continuous_obs_start = as.Date('1920-10-01')
barrow_monthlyavgs = ddply(barrow[barrow$obs_date>=continuous_obs_start,],c('year','month'),function(d){
  data.frame(TMIN=mean(d$TMIN,na.rm=TRUE)
             ,TMAX=mean(d$TMAX,na.rm=TRUE)
             ,PRCP=mean(d$PRCP,na.rm=TRUE)
             ,AWND=mean(d$AWND,na.rm=TRUE)
             )
})
ggplot(barrow_monthlyavgs,aes(year,TMIN))+geom_line()+theme_bw()+facet_wrap(~month)+
  geom_smooth(se=TRUE,color='red',span=0.3)+
  geom_hline(lty=2,aes(yintercept=32))
```

We can see some increase in the monthly average lows for all months, but the Fall and Winter months in particular show a dramatic rise in the last decade. 

October in particular has had no average lows below 15F since 2001, before which average October lows below 15F occurred 72% of the years.

```{r}
mean(subset(barrow_monthlyavgs,month==10 & year<=2001)$TMIN<15)
ggplot(subset(barrow_monthlyavgs,month==10),aes(year,TMIN))+geom_line()+theme_bw()+facet_wrap(~month)+
  geom_smooth(se=TRUE,color='red',span=0.3)+geom_point()+
  geom_hline(lty=2,aes(yintercept=32))
```

Below is some statistical testing on the lows since 2002 showing that this increase is significant, though less pronounced in the summer months than in the fall-winter.

```{r}
summary(aov(TMIN~I(year>2001)*factor(month),data=barrow))
summary(lm(TMIN~I(year>2001)*factor(month),data=barrow))
```
