---
title: "Barrow climate data"
output:
  html_document:
    df_print: paged
---

This is a small analysis of Barrow daily weather summaries sourced from NOAA (https://www.ncdc.noaa.gov/cdo-web/datasets). 

I was inspired to look at Barrow weather data by reading in the Washington Post about how the data there was setting off outlier detection climate models (https://www.washingtonpost.com/news/capital-weather-gang/wp/2017/12/12/barrow-is-warming-so-fast-algorithms-removed-the-data-because-it-seemed-unreal/?utm_term=.7a5b322526b4) 

and reading this climate.gov article about rising average temperatures for October (https://www.climate.gov/news-features/understanding-climate/barrow-alaska-climate-change-action)

```{r}
library(ggplot2)
library(plyr)
library(forecast)
library(ggmap)

# read NOAA daily summary data for Barrow
barrow <- read.csv("~/Documents/Climate-Weather/barrowAK_daily.csv", stringsAsFactors=FALSE)
barrow = barrow[,c('STATION','NAME','LATITUDE','LONGITUDE','ELEVATION','DATE'
                   ,'AWND','PGTM','PRCP','SNOW','TMIN','TAVG','TMAX')]
barrow$obs_date = as.Date(barrow$DATE)
barrow$year = as.numeric(unlist(lapply(barrow$DATE,FUN=function(s){strsplit(s,split='-')[[1]][1]})))
barrow$month = as.numeric(unlist(lapply(barrow$DATE,FUN=function(s){strsplit(s,split='-')[[1]][2]})))
barrow$dayofmonth = as.numeric(unlist(lapply(barrow$DATE,FUN=function(s){strsplit(s,split='-')[[1]][3]})))

str(barrow)
range(barrow$DATE)
```

This data is from the Barrow airport weather station and spans from Sept 1901 to Dec 2017. 

Following the climate.gov article, let's look at monthly averages over time starting in Oct 1920:

```{r}
continuous_obs_start = as.Date('1920-10-01')
barrow_monthlyavgs = ddply(barrow[barrow$obs_date>=continuous_obs_start,],c('year','month'),function(d){
  data.frame(TMIN=mean(d$TMIN,na.rm=TRUE)
             ,TMAX=mean(d$TMAX,na.rm=TRUE)
             ,PRCP=mean(d$PRCP,na.rm=TRUE)
             ,AWND=mean(d$AWND,na.rm=TRUE)
             )
})
ggplot(barrow_monthlyavgs,aes(year,TMIN))+geom_line()+theme_bw()+facet_wrap(~month)+
  geom_smooth(se=TRUE,color='red',span=0.2,method=loess)+
  geom_hline(lty=2,aes(yintercept=32))
```


