---
title: "Portland: frost dates"
output:
  github_document
---

# Study frost dates in Portland OR

For garden planning- for lots of vegetables we should wait until after the last frost of winter to plant. 

According to http://www.multnomahmastergardeners.org/sgp-frostdates, for Portland - Airport:

- Latest Last frost date is April 21
- Average last frost date is March 28
- Earliest last frost date is February 29 (Leap day!?)


## I will take a look at the climate record for Portland airport data.

Data from NOAA: https://w2.weather.gov/climate/local_data.php?wfo=pqr
Daily climate record for Portland OR from airport station, 1940-2019


```{r}
library(dplyr)
library(ggplot2)
library(ggridges)
library(readr)
library(tidyr)
library(tibble)
library(lubridate)
library(forcats)

theme_set(theme_minimal())
```


## Read data in 

```{r}
pdx_climate_raw <- read_csv(
    "Portland_dailyclimatedata.csv", 
    skip = 6)
```

```{r}
colnames(pdx_climate_raw) <- c('year','month','obs_type_abbr',
                               paste("day",as.character(seq(1:31)),sep='_'),
                               'avg_or_total')
```


This data is arranged with columns for the day of month and rows for year, month & observation type:

```{r}
head(pdx_climate_raw)
```

So, we need to do some wrangling to make this data easier to analyze


## initial processing and tidying

This data is formatted with a row per year, month observations for Max Temp, Min Temp, Precipitation, and Snowfall and a column for each day of the month



Notes before getting into this data:
- Plan to pivot this to a longer format so we have a row per date observation and separate avg/totals
- There are missing values from before the first observations in the record indicated by "M"
- We also have NA values for 2020; we can just drop these rows
- We have values of "T" and "T/A" for precip. and snow, meaning trace amounts. We will need to do something with those, probably replace with a value less than the smallest observation above zero, which seems to be 0.01

#### Drop 2020: all observations are missing 

```{r}
pdx_climate_wide0 <- filter(pdx_climate_raw,year<2020)
```


## Daily data processing

### Data reshaping

#### Pivot days columns to longer format

```{r}
pdx_longer <- pdx_climate_wide0 %>% 
  select(!avg_or_total) %>%
  pivot_longer(
    cols = day_1:day_31,
    names_to = 'day'
  ) %>%
  mutate(
    day = as.numeric(substr(day,5,6))
  )

head(pdx_longer)
```


#### drop missing values "M" and "-" 
M is at the beginning of the record
- is used for days that don't exist for that month, like Feb 30
```{r}
pdx_longer <- pdx_longer %>% 
  filter(
    !(value %in% c('M','-'))  # Drop missing records
  )

```

#### Use better names for obs_type

```{r}
obs_type_lookup <- tribble(
  ~obs_type_abbr, ~obs_type,
  "TX", "max_temp",
  "TN", "min_temp",
  "PR", "precip",
  "SN", "snow"
)
```

```{r}
pdx_longer <- pdx_longer %>%
  inner_join(obs_type_lookup, by='obs_type_abbr')
```


#### pivot obs_type wider to new fields


```{r}
pdx_daily <- pdx_longer %>%
  select(year,month,day,obs_type,value) %>%
  pivot_wider(
    id_cols = c(year,month,day),
    names_from = obs_type,
    values_from = value
  )

head(pdx_daily)
```
#### Add date column and day of year

```{r}
pdx_daily <- pdx_daily %>%
  mutate(
    date = as.Date(paste(year,month,day,sep='-')),
  ) %>%
  mutate(
    dayofyr = as.integer(
      date - as.Date(paste(year,'01','01',sep='-'))
    ) + 1
  )
```



# Calculate last frost & first frost dates 

"Frost" refers to a day with min temp <= 32 degrees F

Last frost refers to the last frost date from the previous winter, so calculating it may get a little tricky...

What we could do is make a new column to divide the year in half and get the max date with temp<32 in first half of the year for the last frost and min date with temp<32 in the second half of the year for the first frost

## Last frost of year
```{r}
last_frost <- pdx_daily %>%
  filter(month <= 6) %>%
  filter(min_temp <= 32) %>%
  group_by(year) %>%
  summarize(
    last_frost_dayofyr = max(dayofyr),
    last_frost_date = max(date)
    ) %>%
  mutate(
    last_frost_md = format(last_frost_date, format="%b-%d")
  )
```

### Plots for last frost day of year
```{r}
ggplot(last_frost,aes(last_frost_dayofyr))+geom_histogram(bins=10)
```

```{r}
avg_last_frost_doy <- round(mean(last_frost$last_frost_dayofyr))

pdx_daily %>%
  filter(dayofyr == avg_last_frost_doy) %>%
  group_by(month,day) %>%
  summarize(n=n()) %>%
  arrange(desc(n)) 
```
 
```{r}
ggplot(last_frost,aes(year,last_frost_dayofyr))+geom_point(alpha=0.5)+geom_smooth()
```
#### distribution by decade
```{r}
last_frost <- last_frost %>% 
  mutate(
    decade = fct_rev(factor(floor(year / 10)*10))
  )
```

```{r plot.length=12}
ggplot(last_frost,aes(x=last_frost_dayofyr,y=decade,fill=stat(x)))+
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01,bandwidth=7)+
  scale_fill_viridis_c(name="day of year", option = "C")
```
From these plots we can see the last frost of the year has been occurring earlier in the year from the 1940s to the 2010s

```{r}
last_frost %>% 
  filter(decade == '2010')
```
```{r}
last_frost %>% 
  filter(decade %in%  c('2000','2010')) %>%
  summarize(
    median_doy = median(last_frost_dayofyr),
    p90_doy = quantile(last_frost_dayofyr,probs=0.9)
  )
```

```{r}
pdx_daily %>%
  filter(dayofyr == 77) %>%
  group_by(month,day) %>%
  summarize(n=n()) %>%
  arrange(desc(n)) 
```

```{r}
pdx_daily %>%
  filter(dayofyr == 95) %>%
  group_by(month,day) %>%
  summarize(n=n()) %>%
  arrange(desc(n)) 
```


It looks unlikely, given the observations in the last two decades, that the last frost will occur later than April 5th
