---
title: "Oregon climate data"
output:
  html_document:
    df_print: paged
---

# Exploring Portland climate data

Data from NOAA: https://w2.weather.gov/climate/local_data.php?wfo=pqr

Daily climate record for Portland OR from airport station, 1940-2019

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(tibble)
library(lubridate)

theme_set(theme_minimal())
```


## Read data in 

```{r}
pdx_climate_raw <- read_csv(
    "Portland_dailyclimatedata.csv", 
    skip = 6)
```

```{r}
colnames(pdx_climate_raw)
```

```{r}
colnames(pdx_climate_raw) <- c('year','month','obs_type_abbr',
                               paste("day",as.character(seq(1:31)),sep='_'),
                               'avg_or_total')

colnames(pdx_climate_raw)
```


```{r}
str(pdx_climate_raw)
```


## initial processing and tidying

This data is formatted with a row per year, month observations for Max Temp, Min Temp, Precipitation, and Snowfall and a column for each day of the month

```{r}
head(pdx_climate_raw)
```

```{r}
tail(pdx_climate_raw)
```


Notes before getting into this data:
- Plan to pivot this to a longer format so we have a row per date observation and separate avg/totals
- There are missing values from before the first observations in the record indicated by "M"
- We also have NA values for 2020; we can just drop these rows
- We have values of "T" and "T/A" for precip. and snow, meaning trace amounts. We will need to do something with those, probably replace with a value less than the smallest observation above zero, which seems to be 0.01

#### Drop 2020: all observations are missing 

```{r}
pdx_climate_wide0 <- filter(pdx_climate_raw,year<2020)
```

## Daily data processing

### Data reshaping

#### Pivot days columns to longer format

```{r}
pdx_longer <- pdx_climate_wide0 %>% 
  select(!avg_or_total) %>%
  pivot_longer(
    cols = day_1:day_31,
    names_to = 'day'
  ) %>%
  mutate(
    day = as.numeric(substr(day,5,6))
  )

head(pdx_longer)
```

#### drop missing values "M" and "-" 
M is at the beginning of the record
- is used for days that don't exist for that month, like Feb 30
```{r}
pdx_longer <- pdx_longer %>% 
  filter(
    !(value %in% c('M','-'))  # Drop missing records
  )

```

#### Use better names for obs_type

```{r}
obs_type_lookup <- tribble(
  ~obs_type_abbr, ~obs_type,
  "TX", "max_temp",
  "TN", "min_temp",
  "PR", "precip",
  "SN", "snow"
)
```

```{r}
pdx_longer <- pdx_longer %>%
  inner_join(obs_type_lookup, by='obs_type_abbr')
```


#### pivot obs_type wider to new fields


```{r}
pdx_daily <- pdx_longer %>%
  select(year,month,day,obs_type,value) %>%
  pivot_wider(
    id_cols = c(year,month,day),
    names_from = obs_type,
    values_from = value
  )

head(pdx_daily)
```


## Replace "trace amount" with a small value

Take a look at values that are not 0 or Trace amount
```{r}
pdx_daily %>% 
  select(precip,snow) %>%
  filter(!(precip %in% c('T','T/A','0')) & !(snow %in% c('T','T/A','0'))) %>%
  mutate(
    precip = as.numeric(precip),
    snow = as.numeric(snow)
  ) %>%
  summary
          
```

Looks like 0.01 is smallest measured value for precipitation, 0.1 for snow

Also- there are a few values in snow with "/A" after a valid number, ex:

```{r}
pdx_daily %>% filter(year == 1953 & month == 5 & day %in% c(1,8))
```
I can't seem to find what that means in the NOAA documentation. We will just remove the "/A" value in snow

```{r}
pdx_daily <- pdx_daily %>%
  mutate(
    snow = sub("/A","",snow)
  )
```


```{r}
pdx_daily %>% filter(year == 1953 & month == 5 & day %in% c(1,8))
```


Now to replace Trace amounts

```{r}
pdx_daily <- pdx_daily %>%
  mutate(
    obs_date = make_date(year,month,day),
    max_temp = as.numeric(max_temp),
    min_temp = as.numeric(min_temp),
    precip = as.numeric(ifelse(precip == "T",0.001,precip)),
    snow = as.numeric(ifelse(snow == "T",0.01,snow))
  ) %>% 
  mutate(
    snow = ifelse(is.na(snow),0,snow),
    dayofyear = as.integer(obs_date - make_date(year,1,1)) + 1,
    days_since_autumn = ifelse(month < 10 | (month == 10 & day < 21),
                               obs_date - make_date(year-1,10,21),
                               obs_date - make_date(year,10,21))
  )
  

head(pdx_daily)
```

# Data visualization

## When does it snow in Portland?


```{r}

ggplot(pdx_daily,aes(days_since_autumn,snow,color=year))+geom_point()
```

```{r}
# snow summary by month, day
snow_day_summary <- pdx_daily %>%
  
  filter(month %in% c(11,12,1,2)) %>%
  group_by(month,day) %>%
  summarise(
    obs = n(),
    snow_record = max(snow),
    snow_sum = sum(snow),
    snow_freq = sum(I(snow>0))
  )
head(snow_day_summary)
```

```{r}
ggplot(snow_day_summary,aes(day,snow_freq/obs))+geom_point()+geom_line()+facet_wrap(~factor(month,labels = c("Nov","Dec","Jan","Feb"),levels=c(11,12,1,2)))
```

```{r}
ggplot(snow_day_summary,aes(day,snow_sum))+geom_point()+geom_line()+facet_wrap(~factor(month,labels = c("Nov","Dec","Jan","Feb"),levels=c(11,12,1,2)))
```

```{r}
# monthly snow totals trend over time
pdx_daily %>%
  filter(month %in% c(11,12,1,2)) %>%
  group_by(year,month) %>%
  summarize(snow_total = sum(snow)) %>%
  ggplot(aes(year,snow_total)) + geom_point() + geom_smooth() +facet_wrap(~factor(month,labels = c("Nov","Dec","Jan","Feb"),levels=c(11,12,1,2)))
```

```{r}
# monthly snow freq trend over time
pdx_daily %>%
  filter(month %in% c(11,12,1,2)) %>%
  group_by(year,month) %>%
  summarize(snow_freq = sum(I(snow>0))) %>%
  ggplot(aes(year,snow_freq)) + geom_point() + geom_smooth() +facet_wrap(~factor(month,levels=c(10,11,12,1,2,3)))
```
